---
title: "bdacars"
output: html_notebook
---

Playing around with models and code.

Setting our working directory and initializing Stan for use:
```{r}
setwd('C:/Users/esatu/Dropbox/Aalto/Bayesian Data Analysis/Project')
Sys.setenv(USE_CXX14 = 1)
library("rstan") # observe startup messages
```
Loading our data:
```{r}
data(mtcars)
```
Inspecting our data:
```{r}
head(mtcars, 7)
```

Let's get to know our variables of interest. We expect the \textit{miles per gallon} to decrease as \textit{weight} increases.
```{r}
plot(mtcars$mpg ~ mtcars$wt)
```
This appears to be the case. How would we model this with Stan?
We build our model into the file linear_cars.stan.

Now to model we run:
```{r}
fit <- stan(file = 'linear_cars.stan',
            data = list(n = nrow(mtcars),
                        x = mtcars$wt,
                        y = mtcars$mpg))
```

Let's see what we got.
```{r}
plot(fit)
```
```{r}
traceplot(fit)
```

```{r}
fit_df <- as.data.frame(fit)

library(ggplot2)

ggplot(fit_df, aes(alpha, beta, color=sigma)) +
  geom_point(alpha = 0.3) +
  geom_density2d(color = "white") +
  scale_color_viridis_c(option = 'inferno') +
  labs(title = 'Stan parameter space')
```
```{r}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  geom_abline(aes(intercept = alpha, slope= beta), fit_df,
              alpha = 0.07, color = 'gray50') +
  geom_abline(slope = mean(fit_df$beta),
              intercept = mean(fit_df$alpha),
              color = "pink", size=1) +
  labs(title = 'Bayesian regression on Motor Trends')
```

